Aligning sequence reads, clone sequences and assembly contigs with BWA-MEM

### 2013-07-21 response_to_aaron.md

The rate of mis-seeding is determined by the chance of finding a 19bp exact
seed on a read. Even if the error rate reaches 10%, there will still be a very
high chance to see a 19bp seed given a 1000bp read. The simulated data set uses
high error rate and evenly distributed errors. It is already harder than
typical human resequencing data. On real Illumina human data, the bwa-mem
seeding should be an even less concern.

I have not tried bwa-mem on PacBio reads, and I doubt it will perform well.
Bwa-mem is not designed with PacBio in mind. Bwa-sw may be more accurate on
PacBio data.

> There is one glaring omission in the comparison set of aligners and that is
> [LAST](http://last.cbrc.jp). LAST can do some clever things such as
> qualityÂ­aware gapped alignment scoring and approximate Bayesian aligned read
> pairing, in addition to long sequence alignment. If there is some reason that
> a comparison to LAST isn't possible or appropriate it should be stated so
> rather than leaving us to wonder about it.

@@

I would love to fix these grammatical errors.

> There seems to be a single mismatch penalty used for the gapped alignment,
> however, some nucleotide substitutions happen much more frequently than
> others. At a minimum the ability to treat transitions differently than
> transversions would be nice, though being able to use a 4x4 substitution
> matrix (and/or estimating it directly from the data) would be ideal.

For typical resequencing projects, especially for human resequencing, the
sequencing error rate is much higher than the rate of variations. It is not
necessary to distinguish transitions and transversions. For cross-species
alignment where variations are frequent, I agree specifying a scoring matrix is
preferred. But even in that case, it will have little effect on mapping
accuracy (more on alignment accuracy).

> No mention is made of whether or how the perÂ­base quality estimates produced
> by sequencing instruments are used during gapped alignment.

@@

Base quality is not used because base quality is frequently inaccurate and it
does not seem to bring a substantial improvement on accuracy. More importantly,
mismappings caused by sequencing errors tend not to be recurrent; mismappings
caused by variants tend to be recurrent, but considering base quality does not
help. Only recurrent mismappings have impact on downstream analyses.

> The value of the "SW rescuing" approach is unclear. The manuscript says the
> 2nd best SW score is recorded Â­Â­ what happens when there are many equally
> good alignments? The alignment with the 2nd best score might be minimally
> different. Or is this referring to the best alignment to a different region
> of the reference sequence? If so, the explanation needs to be improved.

@@

> "is more performant" ­> performs better

It will be changed.

####Review of the software

> ...<br> accuracy was evaluated using a short perl script (see file
> bwa\_mem\_eval.tar.bz2 in figshare):
>
>> ./accuracy.pl cbot\_mem.sam < cbot\_mem.both.aln<br> precision:
>> 0.992127563556135 recall: 0.998284561049445<br> ./accuracy.pl
>> lastal\_pairs.sam < cbot\_mem.both.aln<br> precision: 1 recall:
>> 0.988837162737148<br>
> 
>
> So lastal has remarkably good precision but aligns slightly fewer of the
> reads with the recommended settings than bwa mem. If we turn these numbers
> into an F1 score (http://en.wikipedia.org/wiki/F1_score)
> we get lastal: 0.98883 bwa­mem: 0.99519 so if you like the balance of
> precision and recall provided by F1, bwa­mem seems like a good choice. For
> things like identifying rare variants, I will probably stick with lastal when
> possible since the occasional misalignments could potentially end up creating a
> lot of rare variants. It might also be possible to improve lastal's recall
> with a lower e­value threshold, I did not explore this.
